When Risk Hits the Road: Does Danger Drive a Return?

### 2024-12-03 paper/paper.qmd
add plots and tables for predictor varibles

# Load required libraries
library(tidygraph)
library(ggraph)

@@

[plot of distribution of theft incidents across different hours]

@@

[plot of distribution of theft incidents across day of week]

#### Premises Type

The **Premises Type** variable serves as a predictor to analyze the contextual settings where motorbike theft incidents occur. This variable helps identify whether certain types of premises, such as residential areas, commercial establishments, or public parking lots, are associated with higher or lower theft risks. Factors like the availability of surveillance, the density of parked motorbikes, or accessibility to offenders may contribute to variations across premises types. @tbl-premises_summary provides a breakdown of the number of theft incidents by premises type, highlighting locations with elevated or reduced risks.

[table counting the number of theft incident happened in different premises eg house]

@@

[table of some examples of neighborhood ID]

@@

The **Road Surface Condition** variable is a categorical predictor that describes the state of the road at the time of an incident, such as dry, wet, ice, or loose snow. This variable helps analyze how varying surface conditions influence motorbike collision risks.  @plot-road_surface_summary illustrates the number of incidents across various road surface conditions, [offering insights into how different surface types contribute to overall risk patterns.]

[barlot displaying the count of incidents across different road surface conditions]

@@

The **Lighting Condition** variable is a categorical predictor that represents the type of lighting at the time of an incident, including conditions such as daylight, darkness, artificial lighting, or dawn. This variable is crucial for assessing how visibility levels and lighting environments influence the likelihood of motorbike thefts and collisions. Low visibility conditions, such as darkness or poor artificial lighting, are often linked to increased risks. @plt-lighting_summary shows the number of collision incident occurs for each lighting condition, where [xxx] has a highest...

[barplot displaying the count of incidents for different lighting conditions]

@@

The Traffic Control variable categorizes the type of traffic management present at the location of each incident. This includes classifications such as "No Control" and "Traffic Signal." These categories allow for an analysis of how different traffic control measures correlate with collision risks. Contributing factors may include the absence of control mechanisms leading to higher collision probabilities or the efficiency of traffic signals in mitigating such risks.

@tbl-traffic_control_summary highlights the number of incidents across various traffic control types, providing insights into which systems are most associated with increased or decreased collision risks.

[table for counting the number of collision for different traffic control]

@@

[table- 5 examples of different time of day]

#### Visibility

@@

### Theft Sub-Index \[add one sentence of validation\]

The theft sub-index approach is based on two key assumptions. First, it assumes that the observed proportions of theft occurrences across categories, such as hour of the day, day of the week, and premises type, accurately represent the overall theft risk. Second, it assumes that these categories contribute independently to the theft risk, meaning the risk associated with one category does not influence or depend on another.

### Composite Risk Index \[Add one sentence of assumption\]

The validation of the composite Risk Index involves two key steps. First, its correlation with observed collision severity will be tested to ensure alignment with real-world risks. Second, a sensitivity analysis will be conducted by testing alternate weightings for the collision and theft components $w_1$ and $w_2$ to evaluate the robustness and reliability of the final Risk Index.

@@

@@

# Load the data
final_risk_data <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))

# Preprocess lighting conditions: Group 'NA' as 'Other' and create lighting condition categories for the full dataset
merged_data$lighting_conditions <- as.character(merged_data$lighting_conditions)  # Ensure it's a character vector

# Convert "NA" (actual value in the data) to "Other"
merged_data$lighting_conditions[merged_data$lighting_conditions == "NA"] <- "Other"

# Group lighting conditions into four categories, including "Other"
merged_data$lighting_condition_group <- case_when(
  merged_data$lighting_conditions %in% c("Dark", "Dark, Artificial") ~ "Low Light (Dark & Artificial)",
  merged_data$lighting_conditions %in% c("Dawn", "Dawn, Artificial") ~ "Dawn and Artificial Light",
  merged_data$lighting_conditions %in% c("Daylight", "Daylight, Artificial") ~ "Bright Daylight and Artificial Light",
  merged_data$lighting_conditions == "Other" ~ "Other Lighting Conditions"
)

# First Plot: Low Light (Dark & Artificial)
plot_dark <- ggplot(merged_data %>% filter(lighting_condition_group == "Low Light (Dark & Artificial)"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("Low Light (Dark & Artificial)" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# Second Plot: Dawn and Artificial Light
plot_dawn <- ggplot(merged_data %>% filter(lighting_condition_group == "Dawn and Artificial Light"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dawn and Artificial Light" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# Third Plot: Bright Daylight and Artificial Light
plot_daylight <- ggplot(merged_data %>% filter(lighting_condition_group == "Bright Daylight and Artificial Light"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Bright Daylight and Artificial Light" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# Fourth Plot: Other Lighting Conditions
plot_other <- ggplot(merged_data %>% filter(lighting_condition_group == "Other Lighting Conditions"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Other Lighting Conditions" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# Display the four plots side by side
plot_dark
plot_dawn
plot_daylight
plot_other

@@

# Load necessary libraries
library(sf)
library(ggplot2)
library(osmdata)
library(dplyr)

>>>>

# Load required libraries

library(tidygraph)
library(ggraph)

@@

```{r}
#| label: plt-hour_summary
#| fig-cap: "Theft Incidents by Hour of the Day"
#| echo: false
#| warning: false
#| message: false
# Load necessary libraries
library(tidyverse)

# Load the cleaned crime data
cleaned_crime_data <- read_csv(here("data", "02-analysis_data", "cleaned_crime_data.csv"))

# Summarize the number of theft incidents by hour
hour_summary <- cleaned_crime_data %>%
  group_by(hour) %>%
  summarize(theft_count = n()) %>%
  arrange(hour)

# Create the plot
ggplot(hour_summary, aes(x = hour, y = theft_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  scale_x_continuous(breaks = 0:23, labels = sprintf("%02d", 0:23)) + # Ensure 24-hour format
  theme_minimal() +
  labs(

    x = "Hour of the Day (24-hour format)",
    y = "Number of Theft Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| label: plot-day_summary
#| fig-cap: "Theft Incidents by Day of the Week"
#| echo: false
#| warning: false
#| message: false

# Summarize the number of theft incidents by day of the week
day_summary <- cleaned_crime_data %>%
  group_by(day_of_week) %>%
  summarize(theft_count = n()) %>%
  arrange(match(day_of_week, c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

# Create the plot with a different color
ggplot(day_summary, aes(x = day_of_week, y = theft_count)) +
  geom_bar(stat = "identity", fill = "tomato") +
  theme_minimal() +
  labs(

    x = "Day of the Week",
    y = "Number of Theft Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```
#### Premises Type

The **Premises Type** variable serves as a predictor to analyze the contextual settings where motorbike theft incidents occur. This variable helps identify whether certain types of premises, such as residential areas, commercial establishments, or public parking lots, are associated with higher or lower theft risks. Factors like the availability of surveillance, the density of parked motorbikes, or accessibility to offenders may contribute to variations across premises types. @plt-premises_summary provides a breakdown of the number of theft incidents by premises type, highlighting locations with elevated or reduced risks.

```{r}
#| label: plt-premises_summary
#| fig-cap: "Theft Incidents by Day of the Week"
#| echo: false
#| warning: false
#| message: false

# Summarize the number of theft incidents by premises type
premises_summary <- cleaned_crime_data %>%
  group_by(premises_type) %>%
  summarize(theft_count = n()) %>%
  arrange(desc(theft_count))  # Sort by the number of incidents

# Create the plot for premises type
ggplot(premises_summary, aes(x = reorder(premises_type, -theft_count), y = theft_count)) +
  geom_bar(stat = "identity", fill = "dodgerblue") +
  theme_minimal() +
  labs(
    title = "Theft Incidents by Premises Type",
    x = "Premises Type",
    y = "Number of Theft Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| label: tbl-neighborhood_summary
#| fig-cap: "10 Examples of Neighborhood Incidents Count"
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(kableExtra)
library(here)
# Load the cleaned traffic data
cleaned_traffic_data <- read_csv(here("data", "02-analysis_data", "cleaned_traffic_data.csv"))

# Summarize the dataset by Neighborhood ID (hood_158)
neighborhood_summary <- cleaned_traffic_data %>%
  group_by(hood_158) %>%
  summarize(
    incident_count = n()  # Count the number of incidents per neighborhood
  ) %>%
  arrange(desc(incident_count))  # Sort neighborhoods by the number of incidents

# Select a few examples (e.g., top 5 and bottom 5 neighborhoods)
example_neighborhoods <- neighborhood_summary %>%
  slice(c(1:5, (n() - 4):n()))  # Top 5 and bottom 5 neighborhoods

# Create the table with a frame
example_neighborhoods %>%
  kable(
    col.names = c("Neighborhood ID", "Incident Count"),
    align = "cc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

@@

The Road Surface Condition variable is a categorical predictor that describes the state of the road at the time of an incident, including categories such as dry, wet, loose snow, slush, ice, packed snow, gravel, spilled liquid, and others. This variable helps analyze how varying surface conditions influence motorbike collision risks. @plot-road_surface_summary illustrates the number of incidents across all these road surface conditions, revealing that the majority of incidents occur on dry roads, followed by wet roads. Other surface conditions, such as snow, ice, or gravel, account for a much smaller proportion of incidents. This distribution suggests that while dry conditions are most common for collisions, likely due to higher traffic volumes, wet and less stable surfaces may pose proportionally greater risks. These insights emphasize the importance of considering all road conditions when designing targeted road safety measures, particularly during adverse weather, to mitigate collision risks effectively.

```{r}
#| label: plot-road_surface_summary
#| fig-cap: "Incidents by Road Surface Condition"
#| echo: false
#| warning: false
#| message: false

# Summarize the number of incidents by road surface condition
road_surface_summary <- cleaned_traffic_data %>%
  group_by(road_conditions) %>%
  summarize(incident_count = n()) %>%
  arrange(desc(incident_count))  # Sort by number of incidents

# Create the bar plot for all road surface conditions
ggplot(road_surface_summary, aes(x = reorder(road_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "seagreen") +
  theme_minimal() +
  labs(
    title = "Incidents by Road Surface Condition",
    x = "Road Surface Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The Lighting Condition variable is a categorical predictor that represents the type of lighting at the time of an incident, including conditions such as daylight, darkness, artificial lighting, or dawn. This variable is crucial for assessing how visibility levels and lighting environments influence the likelihood of motorbike thefts and collisions. The bar plot @plt-lighting_summary shows that the majority of incidents occur during daylight conditions, which likely reflects higher traffic volumes during daytime hours. Incidents under "Dark" and "Dark, artificial" lighting conditions are significantly fewer but still notable, emphasizing the increased risks associated with reduced visibility. Other conditions, such as "Dawn" and "Dusk," contribute minimally to the total incidents. These insights highlight the importance of targeted safety interventions, such as improved artificial lighting and driver awareness during nighttime and transitional lighting conditions, to mitigate collision risks.
```{r}
#| label: plt-lighting_summary
#| fig-cap: "Incidents by Lighting Conditionk"
#| echo: false
#| warning: false
#| message: false

# Combine "NA" with "Dawn, artificial" and summarize
lighting_summary <- cleaned_traffic_data %>%
  mutate(
    lighting_conditions = case_when(
      lighting_conditions %in% c("NA", "Dawn, artificial") ~ "Dawn, artificial",
      TRUE ~ lighting_conditions
    )
  ) %>%
  filter(lighting_conditions != "Other") %>%  # Exclude "Other"
  group_by(lighting_conditions) %>%
  summarize(incident_count = n()) %>%
  arrange(desc(incident_count))  # Sort by the number of incidents

# Create the updated bar plot
ggplot(lighting_summary, aes(x = reorder(lighting_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  theme_minimal() +
  labs(
    x = "Lighting Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The Traffic Control variable categorizes the type of traffic management present at the location of each incident, such as "No Control," "Traffic Signal," or "Stop Sign." This variable enables an analysis of how different traffic control measures correlate with collision risks. As shown in @tbl-traffic_control_summary, the majority of incidents occur in areas with "No Control" (9,021 incidents), followed closely by areas with "Traffic Signal" (8,035 incidents). This suggests that the absence of traffic control mechanisms, as well as high traffic volumes at signalized intersections, may contribute significantly to collision probabilities. Less frequent traffic control types, such as "Stop Signs" (1,464 incidents) or "Pedestrian Crossovers" (208 incidents), are associated with considerably fewer incidents. These findings underscore the importance of implementing and optimizing traffic control measures, particularly in areas without existing controls, to reduce collision risks effectively.
```{r}
#| label: tbl-traffic_control_summary
#| fig-cap: "Summary of Traffic Control"
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(kableExtra)

# Summarize the dataset by Traffic Control Type
traffic_control_summary <- cleaned_traffic_data %>%
  group_by(traffic_control) %>%
  summarize(
    incident_count = n()  # Count the number of incidents per traffic control type
  ) %>%
  arrange(desc(incident_count))  # Sort by the number of incidents

# Create the table with a frame
traffic_control_summary %>%
  kable(
    col.names = c("Traffic Control", "Incident Count"),
    align = "cc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

@@

```{r}
#| label: tbl-time_of_day_summary
#| fig-cap: "5 Examples of Different Times of Day"
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(kableExtra)

# Select 5 random examples of time_of_day
time_of_day_examples <- cleaned_traffic_data %>%
  select(time_of_day) %>%
  distinct() %>%               # Ensure unique times
  sample_n(5)                  # Randomly select 5 examples

# Create the table with a frame
time_of_day_examples %>%
  kable(
    col.names = c("Time of Day"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

#### Visibility
The Visibility variable is a categorical predictor that describes the environmental visibility conditions at the time of a collision, such as clear, foggy, or reduced due to rain or snow. This variable is critical for assessing how different visibility levels impact the likelihood of collisions. Poor visibility conditions, such as heavy fog, snow, or rain, can obscure drivers’ ability to perceive road hazards or other vehicles, increasing the probability of incidents. Conversely, clear visibility conditions often correlate with lower collision risks.

@plot-visibility_summary provides a barplot summarizing the number of collision incidents under each visibility condition, illustrating how varying environmental factors contribute to changes in collision probabilities. The plot shows that the majority of incidents occur under clear conditions, with a significantly higher incident count than other conditions. Rain, the second most frequent condition, accounts for a smaller but notable proportion of incidents, while other conditions like snow, fog, and strong winds contribute minimally. These findings emphasize the need for proactive measures such as improving road drainage and visibility aids during adverse weather to mitigate collision risks effectively.

```{r}
#| label: plot-visibility_summary
#| fig-cap: "Incidents by Visibility Condition"
#| echo: false
#| warning: false
#| message: false

# Combine "Strong Wind" and "NA" into "Strong Wind" and summarize
visibility_summary <- cleaned_traffic_data %>%
  mutate(
    visibility_conditions = case_when(
      visibility_conditions %in% c("NA", "Strong wind") ~ "Strong wind",
      TRUE ~ visibility_conditions
    )
  ) %>%
  group_by(visibility_conditions) %>%
  summarize(incident_count = n()) %>%
  arrange(desc(incident_count))  # Sort by the number of incidents

# Create the updated bar plot for visibility conditions
ggplot(visibility_summary, aes(x = reorder(visibility_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  theme_minimal() +
  labs(
    x = "Visibility Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

### Theft Sub-Index

The theft sub-index approach is based on two key assumptions. First, it assumes that the observed proportions of theft occurrences across categories, such as hour of the day, day of the week, and premises type, accurately represent the overall theft risk. Second, it assumes that these categories contribute independently to the theft risk, meaning the risk associated with one category does not influence or depend on another.To validate this approach, the calculated values of the theft sub-index were compared against historical crime data trends, confirming that the observed proportions align with real-world patterns of theft distribution across temporal and spatial categories.

### Composite Risk Index

For the assumption of the composite risk index, it is assumed that the weights assigned to the collision probability and theft component reflect their relative importance in contributing to the overall risk, based on the severity and frequency of these events in real-world scenarios. The validation of the composite Risk Index involves two key steps. First, its correlation with observed collision severity will be tested to ensure alignment with real-world risks. Second, a sensitivity analysis will be conducted by testing alternate weightings for the collision and theft components:

$$
w_1 \text{ and } w_2
$$

to evaluate the robustness and reliability of the final Risk Index.

@@

To better understand how environmental factors influence the Risk Index, we examine the distribution of the Risk Index under various road surface conditions. This analysis explores whether different surface types—such as dry, wet, icy, and others—affect the level of risk associated with traffic incidents. The following violin plot visually compares the Risk Index across different road surface conditions, providing insight into how these factors may contribute to the overall risk.
```{r}
#| label: fig-risk-distribution-roadcondition
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Distribution by Road Surface Conditions"
#| fig-pos: H
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Combine "Spilled Liquid" with "Wet" and filter the data
violin_data <- cleaned_traffic_data %>%
  mutate(
    road_conditions = case_when(
      road_conditions == "Spilled liquid" ~ "Wet",
      TRUE ~ road_conditions
    )
  ) %>%
  filter(!is.na(road_conditions), !is.na(risk_index))  # Ensure no NA values

# Create the updated violin plot
ggplot(violin_data, aes(x = road_conditions, y = risk_index, fill = road_conditions)) +
  geom_violin(trim = FALSE) +  # Add violin layers
  scale_fill_brewer(palette = "Set3") +  # Use a color palette for better distinction
  theme_minimal() +
  labs(
    x = "Road Surface Condition",
    y = "Risk Index",
    fill = "Road Conditions"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@fig-risk-distribution-roadcondition reveals notable differences in the Risk Index distribution across various road surface conditions. Wet and icy conditions, in particular, show higher variability in the Risk Index, suggesting that these surfaces may lead to more unpredictable and elevated risk levels. Dry conditions, on the other hand, generally exhibit a lower and more consistent risk. These findings highlight the significant impact that road surface conditions can have on traffic-related risks, underscoring the importance of considering these factors in risk management and safety measures.

@@

# Group lighting conditions into four categories
merged_data$lighting_condition_group <- case_when(
  merged_data$lighting_conditions == "Dark" ~ "Dark",
  merged_data$lighting_conditions == "Daylight" ~ "Daylight",
  merged_data$lighting_conditions == "Dawn" ~ "Dawn",
  merged_data$lighting_conditions %in% c("Dark, artificial", "Daylight, artificial", "Dawn, artificial") ~ "Artificial Lighting",
  TRUE ~ "Other"  # Optional: Handle any remaining uncategorized conditions
)

# First Plot: Dark
plot_dark <- ggplot(merged_data %>% filter(lighting_condition_group == "Dark"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dark" = "#54350a")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Dark Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Second Plot: Daylight
plot_daylight <- ggplot(merged_data %>% filter(lighting_condition_group == "Daylight"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Daylight" = "#FFA07A")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Daylight Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Third Plot: Dawn
plot_dawn <- ggplot(merged_data %>% filter(lighting_condition_group == "Dawn"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dawn" = "#87CEFA")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Dawn Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Fourth Plot: Artificial Lighting
plot_artificial <- ggplot(merged_data %>% filter(lighting_condition_group == "Artificial Lighting"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Artificial Lighting" = "#A9A9A9")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Artificial Lighting Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Display the four plots side by side
plot_dark
plot_daylight
plot_dawn
plot_artificial

@@

# Load necessary libraries
library(sf)
library(ggplot2)

library(osmdata)
library(dplyr)
